\chapter{Publications}
Boudjeniba, Cheïma, Perrine Soret, Diana Trutschel, Antoine Hamon, Valentin Baloche, Bastien Chassagnol, Emiko Desvaux, et al. ‘Consensus Gene Modules Strategy Identifies Candidate Blood-Based Biomarkers for Primary Sjögren’s Disease’. \emph{medRxiv}, 6 July 2023. 
\url{https://doi.org/10.1101/2023.07.05.23292036} \newline

Chassagnol, Bastien, Bichat Antoine, Boudjeniba Cheima, Wuillemin Pierre-Henri, Nuel Grégory, and Becht Etienne. ‘Gaussian Mixtures in R’. \emph{R Journal}, 2023. \newline


Desvaux, Emiko, Antoine Hamon, Sandra Hubert, Cheïma Boudjeniba, Bastien Chassagnol, Jack Swindle, Audrey Aussy, et al. ‘Network-Based Repurposing Identifies Anti-Alarmins as Drug Candidates to Control Severe Lung Inflammation in COVID-19’. \emph{PLOS ONE} 16, no. 7 (juil 2021), 
\url{https://doi.org/10.1371/journal.pone.0254374}. \newline


Soret, Perrine, Christelle Le Dantec, Emiko Desvaux, Nathan Foulquier, Bastien Chassagnol, Sandra Hubert, Christophe Jamin, et al. ‘A New Molecular Classification to Drive Precision Treatment Strategies in Primary Sjögren’s Syndrome’. \emph{Nature Communications 12}, no. 1 (10 June 2021): 3523. 
\url{https://doi.org/10.1038/s41467-021-23472-7}. \newline

\todo{Add archiv link to my most recent decovart algorithm}


\newpage
\chapter{Oral communications}

\section{Peer-reviewed communications}
Bastien, Chassagnol, Nuel Gregory, and Becht Etienne. ‘DeCovarT: Robust deconvolution of cell mixture in transcriptomic samples by leveraging cross-talk between genes, (sciencesconf.org:jds2023:463226)’. Talk of 15 minutes presented at the Journées de Statistique 2023, \emph{54ems Journées de Statistique de la SFdS}, Université libre de Bruxelles, Belgique, 7 July 2023. 
\url{https://drive.google.com/file/d/15cWwt7daYMu6ppwnakWyDNTeaw80tm7I/view?pli=1}.\newline


Bastien, Chassagnol, and Becht Etienne. ‘DeCovarT, a Holistic Deconvolution Algorithm to Explore Highly Heterogenous Tissues’. PhD in 3 minutes presented at the \emph{Seedpods Day} 2023, Servier Paris-Saclay \acrshort{RD} Institute, 29 June 2023. 
\url{https://servier.com/en/newsroom/seedpods-day-boosting-innovation-post-doc-phd-researchers}
\newline


Bastien, Chassagnol, Nuel Gregory, and Becht Etienne. ‘DeCovarT: Robust deconvolution of cell mixture in transcriptomic samples by leveraging cross-talk between genes, (sciencesconf.org:jobim2023:465083) Prix de la meilleure communication orale.’ Talk of 15 minutes presented at the \emph{JOBIM 2023}, Online, 27 June 2023.
 \url{https://www.youtube.com/@sfbisocietefrancaisedebioi5270/videos}
\newline


Bastien, Chassagnol, Nuel Gregory, and Becht Etienne. ‘DeCovarT, a R Package for a Robust Deconvolution of Cell Mixture in Transcriptomic Samples Using a Multivariate Gaussian Generative Framework, Sciencesconf.Org:Rr2023:468014’. Lightning, short talk of 5 minutes presented at the \emph{Rencontres R}, Avignon Université, 21 June 2023.
\url{https://github.com/Rencontres-R/Rencontres_R_2023/blob/main/Presentations/1_Mercredi/3_Lightning_I/1_rencontresR_bastien_chassagnol_short_talk.pdf}\newline


Bastien, Chassagnol, Bichat Antoine, Nuel Gregory, and Becht Etienne. ‘Clustering-Based Models in R, with Application on Univariate Gaussian Mixtures: Review, Evaluation and Extensions’. Talk of 15 minutes presented at the \emph{Rencontres R}, Paris, AgroParisTech, 13 July 2021. 
\url{https://rr2021.sciencesconf.org/356189}

\newpage

\section{Invited seminars}

Bastien, Chassagnol, Becht Etienne, and Nuel Gregory. ‘Robust deconvolution of transcriptomic samples using the gene covariance structure.’ Highlight of 20 minutes presented at the \emph{StatOmique 2022}, Agro Paris-Saclay, 22 place de l’Agronomie à Palaiseau, 25 October 2022. 
\url{https://statomique.netlify.app/media/StatOmique2022-presBChassagnol.pdf}\newline



Bastien, Chassagnol, Yufei, Luo, and Becht Etienne. ‘Robust deconvolution of transcriptomic samples using the gene covariance structure’. Talk of 20 minutes presented at the Conférence \emph{IA et santé}, Centre Henri Lebesgue, Nantes, 29 June 2022. \url{https://www.lebesgue.fr/fr/node/4708}\newline

\section{Poster sessions}

Bastien, Chassagnol, Yufei, Luo, Nuel Gregory, and Becht Etienne. ‘Overview of DeCovarT, a Holistic Method for the Deconvolution of Heterogeneous Transcriptomic Samples, Obtention d’une Bourse Du GDR BIM’. Poster presented at the \emph{ISMB-ECCB 2023, The 31st Annual  Intelligent Systems For Molecular Biology and the 22nd Annual European Conference on Computational Biology}, Online and palais des Congrès, Lyon, France, 24 July 2023.
 \url{https://iscb.junolive.co/ismb2023/live/exhibitor/ismbeccb2023_poster_1420}\newline


Bastien, Chassagnol, Nuel Gregory, and Becht Etienne. ‘Robust deconvolution of transcriptomic samples using the gene covariance structure’. Poster presented at the \emph{JOBIM 2022}, Université de Rennes 1, France, 5 July 2022. 
\url{https://jobim2022.sciencesconf.org/resource}
